---
title: "Summary"
author: "David T. Callaghan"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set( echo=FALSE, message=FALSE, warning=FALSE)
library(printr)
library(knitr)
library(dplyr)
load("SpawnClust.RData")

```

## Cluster Frequency Over Time

I created a frequency per individual plot of spawning clusters which nicely scaled the data among years. Each year has a nice spawning bump in the middle of September.

```{r spawn_time, fig.height=9, fig.width= 8.5}



spawn.num.12<-spawn.num[spawn.num$YEAR==2012,]
spawn.num.13<-spawn.num[spawn.num$YEAR==2013,]
spawn.num.14<-spawn.num[spawn.num$YEAR==2014,]


peakdates<-function(date,freq){
  data<-data.frame(DATE=date,
                    FREQ=freq)
  peak<-data.frame(START=min(data$DATE[data$FREQ>=(max(data$FREQ)/2)]),
                   END=max(data$DATE[data$FREQ>=(max(data$FREQ)/2)]))
  
}

peakdates.12<-peakdates(date=spawn.num.12$DATE,freq=spawn.num.12$FREQ)
peakdates.13<-peakdates(date=spawn.num.13$DATE,freq=spawn.num.13$FREQ)
peakdates.g13<-peakdates(date=spawn.num.13$DATE,freq=spawn.num.13$FREQ13)
peakdates.14<-peakdates(date=spawn.num.14$DATE,freq=spawn.num.14$FREQ) 
peakdates.g14<-peakdates(date=spawn.num.14$DATE,freq=spawn.num.14$FREQ14)

par(mfrow=c(3,1))

#2012
plot(FREQ/24~DATE,data=spawn.num.12,ylim=c(0,3), main="2012",ylab="Freq per Fish",xlab="",
     xlim=as.POSIXct(c("2012-08-15","2012-10-15"),format="%Y-%m-%d", tz="MST"),type="h",col="grey",
     lwd=6,lend=2)
lines(c(peakdates.12$START-12*60*60,peakdates.12$START-12*60*60),c(-1,3.5))
lines(c(peakdates.12$END+12*60*60,peakdates.12$END+12*60*60),c(-1,3.5))

#2013 group 2013
plot(FREQ13/9~DATE,data=spawn.num.13,ylim=c(0,3), main="2013", ylab="Freq per Fish",xlab="",
     xlim=as.POSIXct(c("2013-08-15","2013-10-15"),format="%Y-%m-%d", tz="MST"), type="h",col="grey",
     lwd=6,lend=2)
lines(c(peakdates.g13$START-12*60*60,peakdates.g13$START-12*60*60),c(-1,3.5))
lines(c(peakdates.g13$END+12*60*60,peakdates.g13$END+12*60*60),c(-1,3.5))

#2014 group 2014
plot(FREQ14/3~DATE,data=spawn.num.14,ylim=c(0,3), main="2014", ylab="Freq per Fish",xlab="",
     xlim=as.POSIXct(c("2014-08-15","2014-10-15"),format="%Y-%m-%d", tz="MST"),type="h",col="grey",
     lwd=6,lend=2)
lines(c(peakdates.g14$START-12*60*60,peakdates.g14$START-12*60*60),c(-1,3))
lines(c(peakdates.g14$END+12*60*60,peakdates.g14$END+12*60*60),c(-1,3))



par(mfrow=c(1,1))


```


##Spawning Shoal Use

Below is a bubble pie plot showing cumulative frequency of use for all years (bubble size) and relative use among years (pie slices). It's a bit messy but fun to look at!

```{r fig.height=9.5, fig.width=8 }


library(mapplots)
library(maptools)
library(scales)


data<-first.clust
xyz<-make.xyz(x=data$CLUST.X,y=data$CLUST.Y,z=data$FREQ,group=data$YEAR)

plot(shore_outline)
  
draw.pie(z=xyz$z, x=xyz$x, y=xyz$y, 
        radius=100, col=c(alpha("#FFEF00", 0.7),alpha("#AEC6CF", 0.7),alpha("#FF6961", 0.7)),
        border=c(alpha("#FFEF00", 1),alpha("#AEC6CF", 1),alpha("#FF6961", 1))) 

text(650800,6954500, "Year",font=2) 
legend.pie(650800,6954300, labels=c("2012","2013","2014"), radius = 100,
           bty="n", col=c(alpha("#FFEF00", 0.7),alpha("#AEC6CF", 0.7),alpha("#FF6961", 0.7)),
           border=c(alpha("#FFEF00", 1),alpha("#AEC6CF", 1),alpha("#FF6961", 1)),
           label.dist=1.7)

text(650800,6954000, "Frequency",font=2) 
legend.bubble(650800,6953800, z=92, round=1, n=1, maxradius=100, bty="n",txt.col="transparent")
text(650950,6953800, "92", cex=0.8) 
legend.bubble(650800,6953625, z=46, round=1, n=1, maxradius=50, bty="n",txt.col="transparent")
text(650900,6953625, "46", cex=0.7) 
legend.bubble(650800,6953525, z=23, round=1, n=1, maxradius=25, bty="n",txt.col="transparent")
text(650875,6953525,"23", cex=0.6) 
legend.bubble(650800,6953462, z=12, round=1, n=1, maxradius=13, bty="n",txt.col="transparent")
text(650862,6953462, "12", cex=0.5)               
```
\newpage
Next I plotted the data from the bubble pie graph and created a sorted horizontal frequency distribution to get a better idea of use among sites.

```{r, fig.height=11, fig.width=8.5}

data<-table(ltspawnclust$YEAR,ltspawnclust$SITE)
data<-data[,order(colSums(data))]
barplot(data,
        axes=FALSE,
        xlim=c(0,100),
        xlab="Total Spawning Clusters Formed",
        ylab="Spawning Sites",
        axisnames=FALSE,
        horiz=TRUE,
        col=c(alpha("#FFEF00", 1),alpha("#AEC6CF", 1),alpha("#FF6961", 1)),
        border=c(alpha("#FFEF00", 1),alpha("#AEC6CF", 1),alpha("#FF6961", 1)),
        legend.text=TRUE,
        args.legend=list( y=145, 
                          x=65, 
                          bty = "n",
                          xpd=TRUE,
                          horiz=TRUE,
                          border="white")
        )
axis(1,pos=0)

  
```
\newpage
Next I created a series of bubble plots for each year where the bubble size represents frequency of use in that year.

```{r, fig.height=9.5 ,fig.width=8}


#determine max frequency
max.all<-max(first.clust$FREQ)

#2012
data<-filter(first.clust,YEAR==2012)
#determine scaling factor
s=max(data$FREQ)/max.all

plot(shore_outline, main=data$YEAR[1], col="grey90",border="grey70")

draw.bubble(z=data$FREQ, x=data$CLUST.X, y=data$CLUST.Y, 
         maxradius=200*s, col=alpha("lightblue", 0.3),pch=19)
draw.bubble(z=data$FREQ, x=data$CLUST.X, y=data$CLUST.Y, 
            maxradius=200*s, col=alpha("blue",0.3))


legend.bubble(650800,6954000, z=max.all, round=1, n=3, maxradius=200, bty="n",pch=19,
              col=alpha("lightblue",0.3), txt.cex = 0.7, txt.col="transparent")
legend.bubble(650800,6954000, z=max.all, round=0, n=3, maxradius=200, bty="n",
              col=alpha("blue",0.3),txt.cex = 0.7, txt.col = alpha("blue",0.3))
```
```{r, fig.height=9.5 ,fig.width=8}


#2013
data<-filter(first.clust,YEAR==2013)
#determine scaling factor
s=max(data$FREQ)/max.all
plot(shore_outline, main=data$YEAR[1], col="grey90",border="grey70")

draw.bubble(z=data$FREQ, x=data$CLUST.X, y=data$CLUST.Y, 
            maxradius=200*s, col=alpha("lightblue", 0.3),pch=19)
draw.bubble(z=data$FREQ, x=data$CLUST.X, y=data$CLUST.Y, 
            maxradius=200*s, col=alpha("blue",0.3))


legend.bubble(650800,6954000, z=max.all, round=1, n=3, maxradius=200, bty="n",pch=19,
              col=alpha("lightblue",0.3), txt.cex = 0.7, txt.col="transparent")
legend.bubble(650800,6954000, z=max.all, round=0, n=3, maxradius=200, bty="n",
              col=alpha("blue",0.3),txt.cex = 0.7, txt.col = alpha("blue",0.3))

```
\newpage
It is a tough to see the frequency bubbles in 2014, but I wanted to maintain the same scale as the previous years.
```{r, fig.height=9.5 ,fig.width=8}
#2014
data<-filter(first.clust,YEAR==2014)
#determine scaling factor
s=max(data$FREQ)/max.all

plot(shore_outline, main=data$YEAR[1], col="grey90",border="grey70")

draw.bubble(z=data$FREQ, x=data$CLUST.X, y=data$CLUST.Y, 
            maxradius=200*s, col=alpha("lightblue", 0.3),pch=19)
draw.bubble(z=data$FREQ, x=data$CLUST.X, y=data$CLUST.Y, 
            maxradius=200*s, col=alpha("blue",0.3))


legend.bubble(650800,6954000, z=max.all, round=1, n=3, maxradius=200, bty="n",pch=19,
              col=alpha("lightblue",0.3), txt.cex = 0.7, txt.col="transparent")
legend.bubble(650800,6954000, z=max.all, round=0, n=3, maxradius=200, bty="n",
              col=alpha("blue",0.3),txt.cex = 0.7, txt.col = alpha("blue",0.3))
```
\newpage

##Basic Lake Trout Movement Analysis

Below is a summary of mean daily displacement, persistence index (closer to 1 means straight line, closer to -1 is circles), mean daily acceleration and thier associated standard deviations for male, female and unkown lake trout over 3 years (2012-2014).
```{r}

data<-spawntable[spawntable$MONTH==9,]
#remove month
data<-data[,-2]

#data$ACCEL.SD[is.nan(data$ACCEL.SD)]<-"-"
#data$ACCEL.MEAN[is.nan(data$ACCEL.MEAN)]<-"-"
kable(data, col.names=c("Year","Sex","n","Daily Displacement","Std. Dev.",
                        "Persistence Index","Std. Dev.",
                        "Accel","Std. Dev."),digits=2,align="c")
```
```{r, echo=FALSE, messages=FALSE, cache=FALSE}

#format scientific notation

require(lubridate)
#select only males or females
data<-daily[daily$MONTH==9 & (daily$SEX=="M" | daily$SEX=="F") & daily$TRANSMITTER!="LT-32",]
data$DAY<-day(as.Date(data$DATE))
data$YEAR<-as.factor(data$YEAR)
data$TRANSMITTER<-as.factor(data$TRANSMITTER)
data$SEX<-as.factor(data$SEX)



############################

#daily displacement

require(nlme)
require(MuMIn)
require(lattice)
require(car)
library(effects)
library(stargazer)
```
#Daily Displacement

Here I show square root transformed daily displacement data (sqrt(STEPLENGTH)) for each lake trout over the course of Septmeber. The different colours represent different years. The trend lines generally show a curvelinear relationship for daily displacement over time.

```{r}
xyplot(sqrt(STEPLENGTH) ~ DAY|TRANSMITTER, groups=YEAR, type=c("p","smooth"), data=data)

```
Next I will run a linear mixed effects model to determine if daily displacment differs between sexes. The model uses individuals nested in year as random intercepts to address repeated measures and AR1 correlation structure to address independence.

The full model is as follows:

$$\sqrt{Daily Displacement} \sim Sex + Day + Day^{2} + Year + random (1|Transmitter:Year)$$ 

Model results shown below indicate no significant difference in daily displacement between sexes but that thier is a significant curvelinear relationship with days as well as among years. The figure indicates how these effects relate to daily displacement---note that it is square root transformed to meet parametric assumptions).

```{r}
#test if random effect better than linear regression
#linear model
gls1<-gls(sqrt(STEPLENGTH)~SEX+DAY+I(DAY^2)+YEAR,data=data)
#mixed effects model
mod1<-lme(sqrt(STEPLENGTH)~SEX+DAY+I(DAY^2)+YEAR, random=~1|TRANSMITTER,data=data, method="REML")

#check variance inflation factor
#vif(mod1)
#centre days
data$cDAY<-data$DAY-mean(data$DAY)

#linear model
gls1<-gls(sqrt(STEPLENGTH)~SEX+cDAY+I(cDAY^2)+YEAR,data=data)
#mixed effects model
mod1<-lme(sqrt(STEPLENGTH)~SEX+cDAY+I(cDAY^2)+YEAR, random=~1|TRANSMITTER,data=data, method="REML")
#re-check variance inflation factor
#vif(mod1)

#test is random effects improves model
#anova(gls1,mod1)#random effect better model,

#acf(residuals(mod1, type="normalized"))#looks like correlation

#add AR1 correlation structure to model
mod2<-lme(sqrt(STEPLENGTH)~SEX+cDAY+I(cDAY^2)+YEAR, random=~1|TRANSMITTER, correlation=corAR1(form=~1|TRANSMITTER/YEAR) ,data=data, method="REML")

#the AR1 model is a better fit then the model without the correlation structure
#anova(mod1,mod2)

#re-check autocorrelation
#acf(residuals(mod2, type="normalized"))#looks pretty good


#re-run model 2 with maximum likelihood (ML)
mod2<-lme(sqrt(STEPLENGTH)~SEX+cDAY+I(cDAY^2)+YEAR, random=~1|TRANSMITTER, correlation=corAR1(form=~1|TRANSMITTER/YEAR) ,data=data, method="ML")

#summary(mod2)


#drop SEX
mod2.a<-update(mod2,.~.-SEX)
#check significance of SEX
#anova(mod2,mod2.a)# looks to be insignificant and can be dropped from full model


#drop DAY
#mod2.b<-update(mod2,.~.-cDAY)
#check significance of DAY
#anova(mod2,mod2.b)# significant!

#drop DAY^2
#mod2.c<-update(mod2,.~.-I(cDAY^2))
#check significance of DAY^2
#anova(mod2,mod2.c)# significant!


#drop YEAR
#mod2.d<-update(mod2,.~.-YEAR)
#Check significance of YEAR
#anova(mod2,mod2.d)# significant difference



#full model
mod.dd<-mod2
best.dd<-mod2.a

kable(Anova(mod.dd,type=2))
#r.squaredGLMM(mod.dd)
```


```{r}
plot(allEffects(mod.dd))
#check assumptions
#heterogeneity
#plot(fitted(mod.dd), residuals(mod.dd, type="normalized"), xlab = "Fitted Values", ylab = "Residuals")
#abline(h = 0, lty = 2)
#lines(smooth.spline(fitted(mod.dd), residuals(mod.dd, type="normalized")),col="red")
#normality
#hist(residuals(mod.dd,type="normalized"))
#independence
#acf(residuals(mod.dd,type="normalized"))


```

#Persistence Index

Here is the same plot as before but for persistence index. No transformation was required.

```{r}

####Persistence Index

#hist(data$PERSINDEX)


#boxplot(PERSINDEX~SEX*YEAR,data=data)
xyplot(PERSINDEX ~ DAY|TRANSMITTER, groups=YEAR, type= c("p","smooth"), data=data)
```

I also ran the same mixed model parameters for persitence index. The results show that there is a significant difference between sexes in terms of persistence index with significant changes among days (note days^2 is not significant indicating a linear day relationship) and years.


```{r}

#test if random effect better than linear regression
#linear model
gls1<-gls(PERSINDEX~SEX+cDAY+I(cDAY^2)+YEAR,data=data)
#mixed effects model
mod1<-lme(PERSINDEX~SEX+cDAY+I(cDAY^2)+YEAR, random=~1|TRANSMITTER,data=data, method="REML")

#check variance inflation factor
#vif(mod1)
#anova(gls1,mod1)#random effect better model,
#l.ratio<-anova(gls1,mod1)[[2,"L.Ratio"]] 
#0.5*(1-pchisq(l.ratio,1))
#check for autocorrelation
#acf(residuals(mod1, type="normalized"))#looks like correlation

#add AR1 correlation structure to model
mod2<-lme(PERSINDEX~SEX+cDAY+I(cDAY^2)+YEAR, random=~1|TRANSMITTER, correlation=corAR1(form=~1|TRANSMITTER/YEAR) ,data=data, method="REML")


#the AR1 model is a better fit then the model without the correlation structure
#anova(mod1,mod2)

#re-check autocorrelation
#acf(residuals(mod2, type="normalized"))#looks pretty good

mod2<-lme(PERSINDEX~SEX+cDAY+I(cDAY^2)+YEAR, random=~1|TRANSMITTER, correlation=corAR1(form=~1|TRANSMITTER/YEAR) ,data=data, method="ML")


#summary(mod2)


#drop SEX
#mod2.a<-update(mod2,.~.-SEX)
#check significance of SEX
#anova(mod2,mod2.a)# significant


#drop DAY
#mod2.b<-update(mod2,.~.-cDAY)
#check significance of DAY
#anova(mod2,mod2.b)# significant!

#drop DAY^2
mod2.c<-update(mod2,.~.-I(cDAY^2))
#check significance of DAY^2
#anova(mod2,mod2.c)# insignificant


#drop YEAR
#mod2.d<-update(mod2,.~.-YEAR)
#Check significance of YEAR
#anova(mod2,mod2.d)# significant difference


#best model

mod.pi<-mod2
best.pi<-mod2.c
#r.squaredGLMM(mod.pi)

kable(Anova(mod.pi,type=2))
```

The figure below illustrates the relationships between parameters and persistence index.

```{r}
plot(allEffects(mod.pi))

#check assumptions
#heterogeneity
#plot(fitted(mod.pi), residuals(mod.pi, type="normalized"), xlab = "Fitted Values", ylab = "Residuals")
#abline(h = 0, lty = 2)
#lines(smooth.spline(fitted(mod.pi), residuals(mod.pi, type="normalized")),col="red")
#normality
#hist(residuals(mod.pi,type="normalized"))
#independence
#acf(residuals(mod.pi,type="normalized"))

```

#Daily Mean Acceleration

Here is the same plot from previously but for daily mean acceleration. This was log trasnformed to meet parametric assumptions. Notice the strong horseshoe relationship---looks to be occuring during peak spawn.

```{r}

###Acceleration

#accelerometers not available in 2012
data.ac<-data[data$YEAR!=2012,]
data.ac$YEAR<-droplevels(data.ac$YEAR)
#remove fish without accelerometers
data.ac<-data.ac[!is.na(data.ac$ACCELMEAN),]

#hist(data.ac$ACCELMEAN)
#log transform for normality
#hist(log(data.ac$ACCELMEAN))
#boxplot(log(ACCELMEAN)~SEX,data.ac)
```
```{r}
xyplot(log(ACCELMEAN) ~ DAY|TRANSMITTER, group=YEAR, type= c("p", "smooth"), data=data.ac)
```

I also ran the same mixed model parameters for daily mean acceleration but need to adjust the random component to transmitter since we only had two years of accelerometer data. The results show that there is a significant difference (although not very convincing...) between sexes in terms of acceration with significant changes among days^2 but not days or years.

```{r}

#test if random effect better than linear regression
#linear model
gls1<-gls(log(ACCELMEAN)~SEX+cDAY+I(cDAY^2)+YEAR,data=data.ac)
#mixed effects model
mod1<-lme(log(ACCELMEAN)~SEX+cDAY+I(cDAY^2)+YEAR, random=~1|TRANSMITTER,data=data.ac, method="REML")

#check variance inflation
#vif(mod1)

#anova(gls1,mod1)#random effect better model, 

#check for autocorrelation
#acf(residuals(mod1, type="normalized"))#looks like correlation

#add AR1 correlation structure to model
mod2<-lme(log(ACCELMEAN)~SEX+cDAY+I(cDAY^2)+YEAR, random=~1|TRANSMITTER, correlation=corAR1(form=~1|TRANSMITTER) ,data=data.ac, method="REML")

#the AR1 model is a better fit then the model without the correlation structure
#anova(mod1,mod2)

#re-check autocorrelation
#acf(residuals(mod2, type="normalized"))#looks pretty good

mod2<-lme(log(ACCELMEAN)~SEX+cDAY+I(cDAY^2)+YEAR, random=~1|TRANSMITTER, correlation=corAR1(form=~1|TRANSMITTER) ,data=data.ac, method="ML")

#test significance of intereaction
#mod2.a<-update(mod2,.~.-SEX)# drop interaction
#anova(mod2,mod2.a)#not significant remove

mod2.b<-update(mod2,.~.-cDAY)# drop interaction
#anova(mod2,mod2.b)#not significant remove

#mod2.c<-update(mod2,.~.-I(cDAY^2))# drop interaction
#anova(mod2,mod2.c)#not significant remove

mod2.d<-update(mod2.b,.~.-YEAR)# drop interaction
#anova(mod2,mod2.d)
#not significant remove
#best model



mod.ac<-mod2
best.ac<-mod2.d
#mod.ac<-update(mod.ac,.~.-YEAR)

#r.squaredGLMM(mod.ac)
kable(Anova(mod.ac,type=2))
#check assumptions
#heterogeneity
#plot(fitted(mod.ac), residuals(mod.ac, type="normalized"), xlab = "Fitted Values", ylab = "Residuals")
#abline(h = 0, lty = 2)
#lines(smooth.spline(fitted(mod.ac), residuals(mod.ac, type="normalized")),col="red")
#normality
#hist(residuals(mod.ac,type="normalized"))
#independence
#acf(residuals(mod.ac,type="normalized"))
```

```{r}
plot(allEffects(mod.ac))

 ```
 
 \newpage

 
A summary of all best models (insignificant predictors dropped) and coefficients is shown below. Note these coefficients have not been back-transformed!

\begin{table}[!htbp] \centering 
  \caption{} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{3}{c}{\textit{Dependent variable:}} \\ 
\cline{2-4} 
\\[-1.8ex] & sqrt(STEPLENGTH) & PERSINDEX & log(ACCELMEAN) \\ 
\\[-1.8ex] & (1) & (2) & (3)\\ 
\hline \\[-1.8ex] 
 SEXM &  & $-$0.147$^{***}$ & 0.245$^{*}$ \\ 
  &  & (0.037) & (0.116) \\ 
  & & & \\ 
 cDAY & 0.832$^{***}$ & 0.006$^{***}$ &  \\ 
  & (0.178) & (0.001) &  \\ 
  & & & \\ 
 I(cDAY$\hat{\mkern6mu}$2) & $-$0.057$^{***}$ &  & $-$0.002$^{***}$ \\ 
  & (0.019) &  & (0.001) \\ 
  & & & \\ 
 YEAR2013 & $-$24.266$^{***}$ & $-$0.052 &  \\ 
  & (4.838) & (0.033) &  \\ 
  & & & \\ 
 YEAR2014 & $-$49.946$^{***}$ & $-$0.180$^{***}$ &  \\ 
  & (5.466) & (0.037) &  \\ 
  & & & \\ 
 Constant & 111.651$^{***}$ & 0.246$^{***}$ & $-$0.915$^{***}$ \\ 
  & (6.147) & (0.041) & (0.094) \\ 
  & & & \\ 
\hline \\[-1.8ex] 
Observations & 668 & 668 & 309 \\ 
Log Likelihood & $-$2,889.275 & 107.961 & $-$63.892 \\ 
Akaike Inf. Crit. & 5,794.551 & $-$199.921 & 139.783 \\ 
Bayesian Inf. Crit. & 5,830.585 & $-$163.887 & 162.183 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{3}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\end{tabular} 
\end{table} 

 
 
 
 


